### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex

In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
for conjunctions of atomic and existentially quantified relational expressions (e.g.,
we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').  Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
\el expressions are easily realizable, but the bisimulation algorithm is more complex.

But let start by formally defining the description languages we are going to use, and the notion of bisimulation.

@@

The previous definition formally introduces the syntax and semantics of the
logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{xxx}).  For completeness we mention them briefly here,
and we refer the reader to~\cite{baad:desc03} for further details:

\begin{center}
\begin{tabular}{lp{6cm}}
$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification\\
$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification\\
$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation
\end{tabular}
\end{center}

@@

We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
the case for the $\alc$ language.  Further details can be found
in~\cite{blac:moda01,kurt:expr99}.

\begin{definition}
Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if it is possible to find a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
\begin{enumerate}
\item $(i,i') \in Z$.
\item if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.
\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\end{enumerate}
\end{definition}

Why are bisimulations relevant for computing referring expressions?
The following result provide us with a crucial characterization:

\begin{theorem}[\cite{blac:moda01}]\label{bisim}
Given a finite model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

>>>>

In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
for conjunctions of atomic and existentially quantified relational expressions (e.g.,
we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').
%Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
%\el expressions are easily realizable, but the bisimulation algorithm is more complex.

Let start by formally defining the description languages we are going to use, and the notion of bisimulation.

@@

The previous definition formally introduces the syntax and semantics of the
logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{sec:related}).  For completeness we mention them briefly here,
and we refer the reader to~\cite{baad:desc03} for further details:

\begin{center}
\begin{tabular}{lp{5.9cm}}
$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification.\\
$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification.\\
$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation.
\end{tabular}
\end{center}

@@

We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
the case for the $\alc$ language,further details can be found
in~\cite{blac:moda01,kurt:expr99}.

\begin{definition}
Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if there is a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
\end{definition}

Why are bisimulations relevant for computing referring expressions?
The following result provide a crucial characterization~\cite{blac:moda01}:

\begin{theorem}\label{bisim}
Given a finite model $\gM$, $i$ and $i'$ are $\alc$-bisimilar in $\gM$
if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex

In computational linguistics, the generating referring expression problem (the GRE
problem) can be intuitively presented as follows: given some information about
a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.

By taking a logic perspective, we can recast the GRE problem as an inference problem.
We can think that the information about the group of individuals is provided in the form
of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
associated with the GRE problem will be to find a formula in $\gL$ such that it
exaclty identifies a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define:
\medskip

@@

Consider the following example.  We have three individuals related as indicated in
Figure~\ref{fig:dale-haddock}d.

@@

We can connect description logics to the problem of generating
referring expressions via the concept of \emph{bisimulation classes}.
A bisimulation class is a maximal subset $C$ of the domain such that
all members of $C$ are pairwise bisimilar.  For the finite models we
deal with in GRE, this amounts to saying that there are certain
concepts that are satisfied by all individuals in $C$ and no
individuals that are not in $C$ (by Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
this means that there exists a unique referring expression for some
individual $i$ iff $i$ is alone in its bisimulation class; and if it
is, we can use one of the formulas that are characteristic for $C$ as
the referring expression.

Representing referring expressions as DL concepts is a very natural
perspective.  The purely conjunctive set of propositional symbols that,
e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
description logic.  Relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
We will discuss this idea further in Section~\ref{xxx}.

@@

There exists very efficient algorithms
for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
These algorithms were originally used for automata minimization: given an
automata $A$ they would return a smaller automata $A'$ which recognized the
same language.  We can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
discuss the $\alc$ case.

@@

\begin{algorithm}[t]
\dontprintsemicolon
\caption{Computes $\mathcal{L}$-bisimulation classes}
\label{algo:bisim-l}
\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
\KwOut{A set \RE of formulas  such that
$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
classes of $\gM$.}

@@

Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
the \alc-bi\-sim\-u\-la\-tion classes for a given model.
Algorithm~\ref{algo:bisim-l} iterates over all propositional and
relational symbols of the signature to construct new formulas until
either all formulas in $\RE$ denote singletons (i.e., there is only one
individual that satisfies them), or no progress has been made in the
previous iteration.  In each iteration, it calls the procedure
add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
\in \RE$ which does not denote a singleton and which is
not equivalent to $\varphi$ and to $\neg \varphi$. In this case it
replace $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
$\RE$.

The $\alc$ algorithm is driven by case distinctions between
concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
negations to the characteristic formulas in $\RE$.  This can be inconvenient
in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{xxx}.

@@

@@

\subsection{Some Examples}

Let's see what this algorithm does on the example shown in
Fig.~\ref{fig:dale-haddock}, which is taken from
\newcite{dale91:_gener_refer_expres_invol_relat}.  In the first loop,
the algorithm will add new concepts $\mathsf{floor}$, $\mathsf{bowl}$,
$\mathsf{cup}$, and $\mathsf{table}$; because adding these concepts
makes $\top$ subsumed, $\C$ will consist of these four concepts after
the loop.  Not all of these concepts are singleton; for instance,
$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
the relations to make the concepts more precise.  After the first
iteration over the relations, we have $\C = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup},
\mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
but we haven't distinguished the cups and tables further, but we can
use the distinction between the bowls to distinguish the cups in the
second iteration.  The result of this is $\C = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{floor}), \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.  At this point, all
concepts except for $\mathsf{table}$ are singleton, and further
iterations don't allow us to refine $\mathsf{table}$; so the algorithm
terminates.  Each of the singleton bisimulation classes is represented
by a characteristic concept; for instance, $\C$ tells us that
$\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
\exists \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we
may refer to $c_2$ as ``the cup in the bowl on the table''.

Compared to the \alc\ algorithm, the \el\ algorithm will generally
compute larger bisimulation classes, and compute singleton
bisimulation classes less frequently, because the concepts it can
build can ``see'' fewer differences between individuals.  For example,
in the model shown in Fig.~\ref{fig:el-vs-alc}, the \alc\ algorithm
will compute the concepts $\{\neg \exists R. (\neg P \sqcap \neg Q),
Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each of
which is singleton), but $a$ and $c$ are \el-bisimilar in this model,
and the only \el-concept that is true in $d$ is $\top$, so the \el\
algorithm can only compute the concepts $\{\top, P, P \sqcap Q\}$.
Furthermore, because the \el\ algorithm maintains a weaker invariant
for the classes, it may consider more of these classes simultaneously.
Prima facie, this means that it has worst-case exponential runtime,
given that the whole domain has an exponential number of subsets.
However, it is possible that a more careful analysis of the
combinatorics of the considered subsets reveal that the algorithm is
actually polynomial; it certainly behaves polynomially in our
evaluations (see below).

>>>>

In computational linguistics, the GRE
problem can be intuitively presented as follows: given some information about
a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
and a grammar, find a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.

By taking a logic perspective, we can recast the GRE problem as an inference problem.
We can assume that the information about the group of individuals is provided in the form
of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
associated with the GRE problem will be to find a formula in $\gL$ such that it
exaclty identifies a given individual $i$.
\medskip

@@

Consider the following example.  We have three individuals related as indicated in
Fig.~\ref{fig:dale-haddock}d.

@@

We can connect description logics to the problem of generating
referring expressions via the concept of \emph{bisimulation classes}.
A bisimulation class is a maximal subset $C$ of the domain such that
all members of $C$ are pairwise bisimilar.  For the finite models we
deal with in GRE, this amounts to saying that all individuals in $C$ and only those satisfy the same formulas (Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
this means that there exists a unique referring expression for some
individual $i$ iff $i$ is alone in its bisimulation class; and if it
is, we can use one of the formulas that are characteristic for $C$ as
the referring expression.

Representing referring expressions as DL concepts is a very natural
perspective.  The purely conjunctive set of propositional symbols that,
e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
description logic.  Relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
We will discuss this idea further in Section~\ref{sec:related}.

@@

There exists very efficient algorithms
for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
These algorithms were originally used for automata minimization (given an
automata $A$ they would return a smaller automata $A'$ which recognized the
same language), but we can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
discuss the $\alc$ case.

@@

\begin{algorithm}[t]
\dontprintsemicolon
\caption{Compute $\mathcal{L}$-bisimulation classes}
\label{algo:bisim-l}
\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
\KwOut{A set \RE of formulas  such that
$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
classes of $\gM$.}

@@

Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
the \alc-bi\-sim\-u\-la\-tion classes for a given model.
Algorithm~\ref{algo:bisim-l} iterates over all propositional and
relational symbols of the signature to construct new formulas until
either all formulas in $\RE$ denote singletons (i.e., there is only one
individual that satisfies them), or no progress has been made in the
previous iteration.  In each iteration, it calls the procedure
add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
\in \RE$ which does not denote a singleton and which is
not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
$\RE$.

The $\alc$ algorithm is driven by case distinctions between
concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
negations to the characteristic formulas in $\RE$.  This can be inconvenient
in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{sec:examples}.

@@

Because the \el\ algorithm maintains a weaker invariant
for the classes, it may consider more of these classes simultaneously.
Prima facie, this means that it has worst-case exponential runtime,
given that the whole domain has an exponential number of subsets.
However, it is possible that a more careful analysis of the
combinatorics of the considered subsets reveal that the algorithm is
actually polynomial; it certainly behaves polynomially in our
evaluations (see below).

@@

\subsection{Some Examples}\label{sec:examples}

We will now see what our algorithms do in some interesting
examples. Consider the model shown in
Fig.~\ref{fig:dale-haddock}a, which is taken from
\newcite{dale91:_gener_refer_expres_invol_relat}, and
let's run the $\el$ algorithm.

In the first loop,
the formulas $\mathsf{floor}$, $\mathsf{bowl}$,
$\mathsf{cup}$, and $\mathsf{table}$ will be added to $\RE$ because they subsume $\top$.  Not all of these formulas denote singletons; for instance,
$|\mathsf{cup}|\gM$ contains two individuals.  So we iterate over
the relations to refine our formulas.  After the first
iteration over the relations, we have $\RE = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup},
\mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
but we haven't distinguished the cups and tables further.

Now we can
use the distinction between the bowls to distinguish the cups in the
second iteration.  The result of this is $\RE = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{floor}), \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.

At this point, all
formulas except $\mathsf{table}$ denote singletons, and further
iterations don't allow us to refine $\mathsf{table}$; so the algorithm
terminates.

Each of the singleton bisimulation classes is represented
by a characteristic formula; for instance, $\RE$ tells us that
$\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
\exists \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we
may refer to $c_2$ as ``the cup in the bowl on the table''.

Compared to the \alc\ algorithm, the \el\ algorithm will generally
compute larger bisimulation classes, and compute singleton
bisimulation classes less frequently, because the concepts it can
build can ``see'' fewer differences between individuals.  For example,
in the model shown in Fig.~\ref{fig:dale-haddock}c, the \alc algorithm
will compute the set $\RE = \{\neg \exists R. (\neg P \sqcap \neg Q),
Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each formula denotes a singleton), but $a$ and $c$ are \el-bisimilar in this model,
and the only \el-concept that is true in $d$ is $\top$, so the \el\
algorithm can only compute the formulas $\{\top, P, P \sqcap Q\}$.
